---
title: "Homework 01 - getting started"
author: "Melinda K. Higgins, PhD."
date: "August 30, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document provides some suggested code and examples to get started on your homework 01.

## Get the data

The following code uses the `tidyverse` packages and workflow. As such these examples load (using the `library()` command) these packages:

* `tidyverse`
* `readxl`
* `forcats`

```{r}
# load tidyverse and readxl packages
library(tidyverse)
library(readxl)

# used read_excel() from the readxl package
dat <- read_excel("Dataset_02_fixq2.xlsx")
```

## do a basic summary

```{r}
# the summary() function gives you some basic summary stats
summary(dat)
```

## get some plots

Using the `tidyverse` workflow, the following code uses "pipes" which is the `%>%` symbol to take the output from one line as the input to the next line. So the code below can be read as "take the dataset called `dat` and send it (pipe it) into the `ggplot()` graphics environment and then add `+` a `geom` layer for a histogram using the `geom_histogram()` function." To learn more about `tidyverse`, see [https://www.tidyverse.org/](https://www.tidyverse.org/). The "R Graphics Cookbook" is another fantastic resource for learning how to use `ggplot()` - see [http://www.cookbook-r.com/Graphs/](http://www.cookbook-r.com/Graphs/). There is a lot of great info and examples online, but you can also purchase the book on Amazon.

### Histogram of Age

```{r}
dat %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = Age), 
                 binwidth = 10)
```

### Bar chart of SES

This plot uses the `geom_bar()` graphics layer. However, you'll notice that the categories still show the numeric codes. We could add labels manually, but it gets tedious to do this everytime.

```{r}
dat %>%
  ggplot() +
  geom_bar(mapping = aes(x = SES))

# you can add an x-axis title and labels for the categories
dat %>%
  ggplot() +
  geom_bar(mapping = aes(x = SES)) +
  scale_x_continuous(name = "Socio Economic Status",
                   breaks = c(1,2,3),
                   labels = c("Low Income","Average Income",
                              "High Income"))
```

## add some factor level info - categorical and ordinal data

We can add the labels for the categorical and ordinal data by changes these variables into "factors". In the `tidyverse` world, the package for this is `forcats`.

```{r}
# load the forcats package
library(forcats)

# create objects for the labels and levels for SES
sesLabels <- c("Low Income","Average Income", "High Income")
sesLevels <- c(1,2,3)

# apply the labels and levels for SES using the factor() function
dat$SES <- factor(dat$SES, 
                  levels = sesLevels, 
                  labels = sesLabels)
```

Now that we've create a factor type variable for SES, we can redo our bar chart but now the labels are automatically applied. We do not have to type them in separately using the `scale_x_continuous()` function in the `ggplot2` package.

```{r}
# redo bar plot
dat %>%
  ggplot() +
  geom_bar(mapping = aes(x = SES))

# another way to do the bar plot
dat %>%
  ggplot(aes(SES)) +
  geom_bar()
```

## adding labels for the 5 LiKert scale questions q1,q2,q3,q4,q5

Using a similar workflow to what we did above for SES, let's do it again for q1.

```{r}
# Likert scale coding for q1-q5
qLevels <- c(1,2,3,4,5) 
qLabels <- c("None of the time","a little of the time","some of the time",
             "a lot of the time","all of the time") 

# apply to q1 - you can redo this for the remaining 4 questions
dat$q1 <- factor(dat$q1, 
                 levels = qLevels, 
                 labels = qLabels)
```

In addition to providing labels in plotting routines, the factor coding also helps with the summary stats.

```{r}
summary(dat$q1)
```

And here is a plot of the q1 responses - notice that there are some missing values, `NA`s. There is a second plot made where these are removed using the `is.na()` function and the actual code uses `!is.na()` which says find the data that are NOT (`!`) missing.

```{r}
dat %>%
  ggplot(aes(q1)) +
  geom_bar()

# remove NAs and replot
dat %>%
  subset(!is.na(q1)) %>%
  ggplot(aes(q1)) +
  geom_bar()

```

